Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-07 main.tex
Minor adjustments and adding a few TODOs.

Training this strategy corresponds to finding mappings of states and histories
to an action and a state. Figures~\ref{} show two of the trained finite state
machines...
% TODO Update sentence above.
% TODO Add diagrams of FSMs.

@@

We conducted a tournament with a large collection of strategies from the Axelrod
library, including some additional parametrized strategies (e.g. various parameter
choices for Generous Tit For Tat). These are listed in
Appendix~\ref{app:list_of_players}.
The top 11 performing strategies by median payoff are all strategies trained to maximize
% TODO Potentially modify this as more repetitions are analysed.
total payoff against a subset of the strategies (Table~\ref{tbl:standard_score}).
The next strategy is Desired Belief Strategy (DBS)
% TODO Add reference for DBS
which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table,
Fool Me Once, a grudging strategy that defects indefinitely on
the second defection, and Omega Tit For Tat \cite{Slany2007}.
All strategies in the tournament follow a simple set of
rules in accordance with earlier tournaments:

@@

% Table of best strategies
\begin{table}[!hbtp]
        \centering
        \input{./assets/standard_top_15_scores_summary.tex}
        \caption{Standard Tournament: Mean score per turn of top 15 strategies
            (ranked by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments).
        The leaderboard is dominated by the machine learning strategies. ~$^{*}$
        indicates that the strategy was trained.}
        \label{tbl:standard_score}
\end{table}

@@

Table~\ref{tbl:standard_wins_top_winners} shows the same information as
Table~\ref{tbl:standard_wins} but for the top 15 strategies who win the most
head to head matches.
% TODO Add more?

@@

Using the method of fingerprinting described in \cite{ashlock2008fingerprinting,
ashlock2009fingerprint}
% TODO This is actually a numerical approximation of the fingerprinting
% described in ashlock's paper. There they consider a fingerprint to be an
% analytic function.
we can compare
strategies. For the top performing noisy strategies there is a striking similarity
in the fingerprints.

@@

    \caption{Comparison of Fingerprints for Standard Tournament Top 3}
    \label{fig:comparison_fingerprint_standard}
\end{figure}
% TODO Add text ref to this

Figure~\ref{fig:comparison_cooperation_heatmaps_standard} shows the rate of
cooperation in each round for the top three strategies.

@@

Table~\ref{tbl:noisy_wins_top_winners} shows the same information as
Table~\ref{tbl:noisy_wins} but for the top 15 strategies who win the most
head to head matches.
% TODO Add more?

@@

Again we compare the fingerprints. For the top performing noisy strategies there
is a striking similarity in the fingerprints which indicates that the strategies
may behave similarly in principle.

@@

    \caption{Comparison of Fingerprints for Noisy (5\%) Tournament Top 3}
    \label{fig:comparison_fingerprint_noisy}
\end{figure}
% TODO Add text ref to this

Figure~\ref{fig:comparison_cooperation_heatmaps_noisy} shows the rate of
cooperation in each round for the top three strategies.

@@

All training code is available on github. There are objective functions for
% TODO We should archive the code.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. A simple evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategies did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well. Single layer neural networks also performed well
though these had some aspect of human involvement in the selection of features.
The success of the other strategy types suggests that a deeper network that
incorporates feature engineering would likely also perform well.

In opposition to historical tournament results and community folklore,
our results show that complex strategies can be very effective for the
IPD. It is not the complexity of strategies that is disadvantageous; rather that directly
designing a broadly effective strategy is no easy task. Of all the human-designed
strategies in the library, only DBS consistently performs well, and it is
substantially more complex than traditional tournament winners like TFT, OmegaTFT,
and zero determinant strategies. Furthermore, dealing with noise is difficult
for most strategies. Two strategies designed specifically to account for noise,
DBS and OmegaTFT, perform well and only DBS performs better than our trained
strategies.

Of the strategies trained to maximize their median score all are generally
cooperative, not defecting until the opponent defects. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable evolutionary equilibrium for the prisoner's
dilemma. Specifically we note that the reinforcement learning process for maximizing
payout does not lead to exploitative zero determinant strategies, which may also
be a result of the collection of training strategies, of which several retaliate harshly.

We take the liberty of generalizing from the results of this study. For the trained
strategies utilizing look up tables we generally found those that incorporate
one of more of the initial rounds of play outperformed those that did not. The
strategies based on neural networks and finite state machines also are able to
condition throughout a match on the first rounds of play. Accordingly, we conclude
that first impressions matter in the IPD\@. The best strategies are nice (never
defecting first) and this property could be further investigated with the library
in future work by e.g. forcing all strategies to defect on the first round.

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained regularly. Most of the strategies were
trained on an earlier version of the library (v2.2.0) that did not include DBS
and several other opponents. The precise parameters that are optimal will depend
on the pool of opponents. Moreover we have not extensively trained strategies to
determine the minimum parameters that are sufficient -- neural networks with
fewer nodes and features and finite state machines with fewer states may suffice.
See \cite{ashlock2013impact} for discussion of resource availability for IPD
strategies. It may be possible to train strategies more effective in noisy tournaments
than DBS.

Future work:
* spatial tournaments and other variants
* Additional strategy archetypes by the Ashlocks, e.g. function stacks, binary
decision players
* further refine features and training parameters

@@

>>>>

Training this strategy corresponds to finding mappings of states and histories
to an action and a state. Figures~\ref{} show two of the trained finite state
machines...
% TODO Update sentence above.
% TODO Add diagrams of FSMs. Note how the strategies are called FSM 16 but as is
% evident in the diagram, it is in fact a 14 state fsm.

@@

We conducted a tournament with a large collection of strategies from the Axelrod
library, including some additional parametrized strategies (e.g. various parameter
choices for Generous Tit For Tat). These are listed in
Appendix~\ref{app:list_of_players}.
The top 11 performing strategies by median payoff are all strategies trained to maximize
% TODO Potentially modify this as more repetitions are analysed.
total payoff against a subset of the strategies (Table~\ref{tbl:standard_score}).
The next strategy is Desired Belief Strategy (DBS) \cite{Au2006}.
which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table,
Fool Me Once, a grudging strategy that defects indefinitely on
the second defection, and Omega Tit For Tat \cite{kendall2007iterated}.
All strategies in the tournament follow a simple set of
rules in accordance with earlier tournaments:

@@

% Table of best strategies
\begin{table}[!hbtp]
        \centering
        \input{./assets/standard_top_15_scores_summary.tex}
        \caption{Standard Tournament: Mean score per turn of top 15 strategies
            (ranked by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments).
        The leaderboard is dominated by the trained strategies. Starred
        strategies $^{*}$
        indicates that the strategy was trained.}
        \label{tbl:standard_score}
\end{table}

@@

Table~\ref{tbl:standard_wins_top_winners} shows the same information as
Table~\ref{tbl:standard_wins} but for the top 15 strategies who win the most
head to head matches.
% TODO Add more discussion?

@@

Using the a numerical method of fingerprinting based on
\cite{ashlock2008fingerprinting, ashlock2009fingerprint} we can compare
strategies. The fingerprints of the top performing standard strategies are shown
in Figures~\ref{fig:comparison_fingerprint_standard}. There is a striking
similarity in the fingerprints.

@@

    \caption{Comparison of Fingerprints for Standard Tournament Top 3}
    \label{fig:comparison_fingerprint_standard}
\end{figure}

Figure~\ref{fig:comparison_cooperation_heatmaps_standard} shows the rate of
cooperation in each round for the top three strategies.
% TODO Discuss more? (If we want to keep this)

@@

Table~\ref{tbl:noisy_wins_top_winners} shows the same information as
Table~\ref{tbl:noisy_wins} but for the top 15 strategies who win the most
head to head matches.
% TODO Add more discussion?

@@

Again we compare the fingerprints. For the top performing noisy strategies
(Figure~\ref{fig:comparison_fingerprint_noisy}) there
is a striking similarity in the fingerprints which indicates that the strategies
may behave similarly in principle.

@@

    \caption{Comparison of Fingerprints for Noisy (5\%) Tournament Top 3}
    \label{fig:comparison_fingerprint_noisy}
\end{figure}

Figure~\ref{fig:comparison_cooperation_heatmaps_noisy} shows the rate of
cooperation in each round for the top three strategies.
% TODO Add more discussion

@@

All training code is available on github. There are objective functions for
% TODO Include pointer when code is archived.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. An evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategies did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well. Single layer neural networks also performed well
though these had some aspect of human involvement in the selection of features.
The success of the other strategy types suggests that a deeper network that
incorporates feature engineering would likely also perform well.

In opposition to historical tournament results and community folklore,
our results show that complex strategies can be very effective for the
IPD\@. It is not the complexity of strategies that is disadvantageous; rather that directly
designing a broadly effective strategy is no easy task. Of all the human-designed
strategies in the library, only DBS consistently performs well, and it is
substantially more complex than traditional tournament winners like TFT, OmegaTFT,
and zero determinant strategies. Furthermore, dealing with noise is difficult
for most strategies. Two strategies designed specifically to account for noise,
DBS and OmegaTFT, perform well and only DBS performs better than the trained
strategies and \textbf{only} in the noisy context.

Of the strategies trained to maximize their average score all are generally
cooperative, not defecting until the opponent defects. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable evolutionary equilibrium for the prisoner's
dilemma. Specifically it is noted that the reinforcement learning process for maximizing
payoff does not lead to exploitative zero determinant strategies, which may also
be a result of the collection of training strategies, of which several retaliate harshly.

We take the liberty of generalizing from the results of this study. For the trained
strategies utilizing look up tables we generally found those that incorporate
one or more of the initial rounds of play outperformed those that did not. The
strategies based on neural networks and finite state machines also are able to
condition throughout a match on the first rounds of play. Accordingly, we conclude
that first impressions matter in the IPD\@. The best strategies are nice (never
defecting first) and this property could be further investigated with the
Axelrod library
in future work by e.g.\ forcing all strategies to defect on the first round.

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained regularly. Most of the strategies were
trained on an earlier version of the library (v2.2.0) that did not include DBS
% Cite v2.2.0
and several other opponents. The precise parameters that are optimal will depend
on the pool of opponents. Moreover we have not extensively trained strategies to
determine the minimum parameters that are sufficient -- neural networks with
fewer nodes and features and finite state machines with fewer states may suffice.
See \cite{ashlock2013impact} for discussion of resource availability for IPD
strategies. It may be possible to train strategies more effective in noisy tournaments
than DBS.

Future work:
% TODO Write the future work section
%* spatial tournaments and other variants
%* Additional strategy archetypes by the Ashlocks, e.g. function stacks, binary
%decision players
%* further refine features and training parameters

@@

The players used for this study are from Axelrod version 2.13.0. The starred $*$
strategies indicate those that have been trained using reinforcement learning
algorithms.
